---
title: "Data Science Salaries 2023/2024"
subtitle: "Reproducible Research Project"
author: "Dilara Ozdil, Eljan Abbaszada, Paula Gwanchele"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    fig_width: 8
    fig_height: 5
  pcleaned_data_document:
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo    = TRUE,
  fig.align = "center"
)
```

# Set working directory to the current folder

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Introduction

This report investigates the determinants of data science salaries using a
cross-sectional dataset of reported compensation. We combine exploratory
data analysis (EDA) with a fixed-effects regression framework to quantify
how experience, work model (remote/hybrid/on-site), company size, and
year affect (log) salary, while absorbing country-level heterogeneity
through location fixed effects. We then assess the robustness of the main
findings to (i) restricting the sample to full-time employees, (ii)
replacing company-location with employee-residence fixed effects, (iii)
trimming the top 1% of salaries, and (iv) allowing the remote-work effect
to vary by experience level.

# Data

## Dependencies

```{r libraries}
required <- c("tidyverse", "fixest", "modelsummary", "scales")
to_install <- setdiff(required, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install)

library(tidyverse)
library(scales)
library(fixest)
library(modelsummary)
```

## Load the raw data

```{r load-data}
raw_data <- read_csv("data_clean.csv")
glimpse(raw_data)
```

## Cleaning and feature construction

A single cleaning pipeline is reused by both the EDA and the econometric
models, removing the duplicated cleaning logic from the two original
scripts.

```{r clean-data}
cleaned_data <- raw_data %>%
  drop_na() %>%
  filter(salary_in_usd > 0) %>%
  mutate(
    experience_level = factor(
      experience_level,
      levels = c("Entry-level", "Mid-level", "Senior-level", "Executive-level")
    ),
    company_size       = factor(company_size, levels = c("Small", "Medium", "Large")),
    work_year          = as.factor(work_year),
    employment_type    = as.factor(employment_type),
    work_models        = as.factor(work_models),
    company_location   = as.factor(company_location),
    employee_residence = as.factor(employee_residence),
    log_salary         = log(salary_in_usd),
    job_category = case_when(
      str_detect(job_title, "Data Scientist")      ~ "Data Science",
      str_detect(job_title, "Data Engineer")       ~ "Data Engineering",
      str_detect(job_title, "Analyst")             ~ "Data Analysis",
      str_detect(job_title, "Machine Learning|ML") ~ "Machine Learning",
      str_detect(job_title, "Manager|Lead|Head")   ~ "Management",
      TRUE                                          ~ "Other"
    )
  ) %>%
  select(-any_of(c("salary", "salary_currency")))

write_csv(cleaned_data, "data_clean.csv")
dim(cleaned_data)
```

# Exploratory Data Analysis

## Experience level vs. salary

```{r eda-experience}
ggplot(cleaned_data, aes(x = experience_level, y = salary_in_usd, fill = experience_level)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Does Experience Actually Pay Off?",
       subtitle = "Salary distribution across experience levels",
       x = "Level of Experience", y = "Salary (USD)") +
  theme_minimal() + theme(legend.position = "none")
```

## Salary distribution

```{r eda-distribution}
ggplot(cleaned_data, aes(x = salary_in_usd)) +
  geom_histogram(fill = "#2d5a3f", color = "white", bins = 30) +
  scale_x_continuous(labels = label_dollar()) +
  labs(title = "The Spread of Data Science Salaries",
       subtitle = "Most salaries cluster between $100k and $200k",
       x = "Salary (USD)", y = "Number of Employees") +
  theme_minimal()
```

## Average salary by job category

```{r eda-job-category}
category_summary <- cleaned_data %>%
  group_by(job_category) %>%
  summarise(avg_salary = mean(salary_in_usd), .groups = "drop") %>%
  arrange(desc(avg_salary))

ggplot(category_summary,
       aes(x = reorder(job_category, avg_salary), y = avg_salary, fill = job_category)) +
  geom_col() + coord_flip() +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Which Field Pays the Best?",
       x = "Job Category", y = "Average Salary (USD)") +
  theme_minimal() + theme(legend.position = "none")
```

## Remote vs. on-site

```{r eda-remote}
ggplot(cleaned_data, aes(x = work_models, y = salary_in_usd, fill = work_models)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, color = "black", outlier.shape = NA) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Remote vs. On-site: Is There a Pay Gap?",
       x = "Work Model", y = "Salary (USD)") +
  theme_minimal() + theme(legend.position = "none")
```

## Company size

```{r eda-company-size}
ggplot(cleaned_data, aes(x = company_size, y = salary_in_usd, fill = company_size)) +
  geom_boxplot() +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Does Company Size Matter?",
       x = "Company Size", y = "Salary (USD)") +
  theme_minimal() + theme(legend.position = "none")
```

## Top 10 locations by average salary

```{r eda-geo}
geo_summary <- cleaned_data %>%
  group_by(company_location) %>%
  filter(n() > 10) %>%
  summarise(avg_salary = mean(salary_in_usd), .groups = "drop") %>%
  arrange(desc(avg_salary)) %>%
  slice_head(n = 10)

ggplot(geo_summary, aes(x = reorder(company_location, avg_salary), y = avg_salary)) +
  geom_col(fill = "#5a8a6e") + coord_flip() +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Top 10 Locations with Highest Average Salaries",
       subtitle = "Only countries with more than 10 reported roles included",
       x = "Country", y = "Average Salary (USD)") +
  theme_minimal()
```

# Model

```{r}
model1 <- feols(
  log_salary ~
    experience_level +
    work_models +
    company_size +
    work_year |
    company_location,
  
  data = cleaned_data,
  
  vcov = "hetero"
)
summary(model1)
```


## Full-time employees only

```{r}
cleaned_data_fulltime <- cleaned_data %>%
  filter(employment_type == "Full-time")

model2 <- feols(
  log_salary ~
    experience_level +
    work_models +
    company_size +
    work_year |
    company_location,
  
  data = cleaned_data_fulltime,
  
  vcov = "hetero"
)
summary(model2)
```

## Employee-residence fixed effects

```{r}
model3 <- feols(
  log_salary ~
    experience_level +
    work_models +
    company_size +
    work_year |
    employee_residence,
  
  data = cleaned_data,
  
  vcov = "hetero"
)
summary(model3)
```

## Trimming the top 1% of salaries

```{r robust-trimmed}
salary_cutoff <- quantile(cleaned_data$salary_in_usd, 0.99)

cleaned_data_trimmed <- cleaned_data %>%
  filter(salary_in_usd < salary_cutoff)

model4 <- feols(
  log_salary ~
    experience_level +
    work_models +
    company_size +
    work_year |
    company_location,
  
  data = cleaned_data_trimmed,
  
  vcov = "hetero"
)
summary(model4)
```

## Heterogeneous effect of remote work by experience

```{r robust-interaction}
model5 <- feols(
  log_salary ~
    experience_level * work_models +
    company_size +
    work_year |
    company_location,
  
  data = cleaned_data,
  
  vcov = "hetero"
)
summary(model5)
```

# Results

The main specification confirms the patterns visible in the EDA:
salaries rise sharply with experience, and there is meaningful variation
across work models and company sizes once country effects are absorbed.
Coefficients are in log points and approximate percentage changes
relative to the omitted category.

## Combined regression table

```{r results-table}
modelsummary(
  list(
    "Main Model"     = model1,
    "Full-Time Only" = model2,
    "Residence FE"   = model3,
    "Trimmed Sample" = model4,
    "Remote x Exp."  = model5
  ),
  stars  = TRUE,
  output = "markdown"
)
```
# Robustness

Columns 2–5 of the table above re-estimate the wage equation under
alternative modelling choices.

- **Full-time only (model 2):** restricting to full-time employees
  removes part-time and contract workers as a source of noise.
- **Employee-residence (model 3):** absorbing the worker's country
  of residence rather than the company's location addresses concerns
  about cross-border remote arrangements.
- **Trimmed sample (model 4):** removing the top 1% of salaries checks
  that extreme top earners are not driving the experience premium.
- **Remote × experience (model 5):** allowing the remote-work effect
  to vary by experience level tests whether the average premium hides
  meaningful heterogeneity.

Across all four alternatives, the experience-level coefficients remain
large, positive, and ordered in the expected direction, and the
work-model and company-size effects keep their sign and rough
magnitude — indicating that the headline findings are not artefacts of
sample composition or fixed-effects choice.

# Conclusion

The combined evidence suggests that experience is the single largest
determinant of compensation in data science roles, with executive-level
employees earning substantially more than entry-level peers even after
absorbing country-level cost-of-living differences. Company size is
positively associated with pay, while the remote-work premium is small
on average but interacts with experience — senior workers benefit
differently from remote arrangements than juniors. Results are stable
across the full-time subsample, an alternative residence-based fixed
effect, and a top-1% trimmed sample, indicating that the headline
findings are not driven by part-time outliers, country mis-coding, or
extreme top earners.

# Session info

```{r session-info}
sessionInfo()
```
